---
title: "Unit5 -- reshaping data with tidyr"
author: "Bruce Thomas"
date: "8/11/2021"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
  slidy_presentation: default
  ioslides_presentation: default
  word_document:
    toc: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message=FALSE)
uid<-rstudioapi::userIdentity()
filePath<-file.path("C:/Users/",uid,"/Navitas Life Sciences/NDS R Training - Documents/General/Curriculum", fsep='/')
print(filePath)
setwd(filePath)

```
## Objectives

### After completing this unit, R Studio users should :
- (1) understand how and when to use TIDYR pivot* fuunctions to reshape data
- (2) know how to remove string data from column names
- (3) know how to remove numeric data from column names
- (4) know how to convert numeric data to integers
- (5) know how to use pivot* functions to handle  multiple observations pre row in a data frame

see https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html for a detailed discussion

***


```{r reshape}

```
***
```{r pivot_string}

```
***
```{r pivot_numeric}

```
***
```{r pivot_numeric_integer}

```
***
```{r pivot_multiple}

```
***