
---
title: "Unit 0: packages and libraries in R"
output:
  word_document:
    toc: yes  
  slidy_presentation: default
  ioslides_presentation: default
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=FALSE,warning=FALSE, message=FALSE)


```

## Objectives

#### After completing this unit, programmers should be able to:

- (1)  understand how RStudio modules/packages are installed and updated.
- (2)  explain the difference between a package and a library
- (3)  use the interface in R studio to add a package
- (4)  install packages from the console window in R Studio
- (5)  Understand when R Studio updates occur.


***
## R Versions

Since R is an open source suite of tools, the various components and their dependencies are packaged and available for installing on a users' PC. as of this writing, the current  version of R is 4.1.0. It is possible to get out ahead of the client with a newer version, and your current code may not work of the client's version does not yet have the functionality or bug fixes installed.  R Studio Enterprise may also provide R packages based on different versions.


```{r "Version"}
## You can get information about the version  of R governing your current R Studio session by typing the following at the command prompt in the R STudio Console window

version
```
***

## R data science packages
The Comprehensive R Archive Network (CRAN) is a large open source repository containing various R packages. Each of these packages consists of one or more files, R scripts and perhaps compiled code and even data.  To investigate, you can enter installed.packages() at the console prompt in RStudio. This function call returns a data structure ( a matrix,or an array ) to the screen. 

```{r "install_package",results='asis'}

## What type -- or class--of object is returned when you invoke the installed.packages function?

class(installed.packages())

## returns dimensions of the array, in Row,Column format
dim(installed.packages())

## or, the installed.packages function can be passed to the print function. 
## look at the 120th installed package, all columns
print(installed.packages()[120,])

##100th package, columns 1,2 and 3
print(installed.packages()[100,1:3])

## Since the detail is extensive, we can work with this new object to find the names of the columns. 
## here, we are looking at the contents of column 1 for all rows.
print(installed.packages()[,1])

#In addition to the print() function, R provides some other functions for looking at what is in an R object. The View() function is particularly useful, it creates a temporary spreadsheet which is navigable inside R Studio. Try changing 'print' to 'View' in the above examples, entering the text at the R Studio Console.

```
## R libraries 

a library function call invokes the package from the group of installed packages available for the R Version for use in an RStudio Session. These packages are installed in the R Studio space reserved for the user on a client or the desktop.   The R version in use determines where these are updated on the user's machine. Once a package is installed, you pass the name to the library() function. If you have installed multiple versions of R, things can get messy. If in doubt, These can be deleted and reinstalled.


Later in this lesson, we will install a few utility packages from CRAN. Before installing any R packages, it is advisable to run the update.packages() function so that the packages are consistent with the version coming from the CRAN repository.

```{r "update_packages"}

## R may issue a message when you try to update packages in MS Windows 10:: 'lib = "C:/Program Files/R/R-4.1.0/library"' is not writable. . The workaround is to navigate to the library in Windows and unset the read-only flag as administrator.
#Enter the following command without the comment '##' at the R STudio console and press Enter. You may be prompted to agree to update each package.

##update.packages("tidyverse",ask=TRUE)

```

***


```{r "install_packages"}

## Create a character vector to pass to the install.packages() function. NAmes should be quoted and are case-sensitive!

#Type the following command without the '##'comment at the R Studio console and press Enter. You may be prompted to agree to update each package. 
## install.packages(c("here","janitor"))

```


***

## Installing packages from the R Studio InterFace

The R Studio IDE provides a Packages pane. It is possible to install packages that way as well. You should try it 
by install.packages("tidyverse"). Tidyverse is a data science analytical package that creates data Structures similar to SAS datasets used in clinical trials.
***

## Libraries


```{r "library"}
## R libraries refer to physical locations on the user's R setup where these packages and their base dependencies are installed. The R version in use determines where these are updated on the user's machine. Once a package is installed, you pass the anme to the library() function so you can use it.
##Load the tidyverse data analysis package. This will form the heart of this training
library (tidyverse)
```


