
---
title: "Unit 0: packages and libraries in R"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
  slidy_presentation: default
  ioslides_presentation: default
  word_document:
    toc: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=FALSE,warning=FALSE, message=FALSE)
```

## Objectives

#### After completing this unit, programmers should be able to:

- (1)  understand how RStudio modules/packages are installed and updated for a 
given version of R.  
- (2)  explain the difference between a package and a library
- (3)  use the interface in R studio to add a package
- (4)  install packages from the console window in R Studio
- (5)  Understand when R Studio updates occur.

***

## R Versions

Since R is an open source suite of tools, the various components and their 
dependencies are packaged and available for installing on a users' PC. as of 
this writing, the current  version of R is 4.1.0. It is possible to get out 
ahead of the client with a newer version, and your current code may not work if
the client's version does not yet have the functionality or bug fixes installed.  
R Studio Enterprise may also provide R packages based on different versions, so 
it is important to verify that the version in use is what the Clinical Trials 
customer is using.


```{r "Version"}
## You can get information about the version  of R governing your current R Studio session by typing the following at the command prompt in the R Studio Console window

version
```
***

## R data science packages

The Comprehensive R Archive Network (CRAN) is a large open source repository 
containing various R packages. Many of these were developed for statistical 
analysis and modeling in a variety of fields ranging from Botany to Epidemiology. 
In general, R packages consist of one or more files, R scripts and perhaps 
compiled code and data designed to help people perform data manipulation and 
analysis. R studio comes with several packages already installed; these are 
usually referred to as the Base packages. While R's basic packages are fairly 
stable, it is frequently necessary to download additional packages to deal with 
data issues and reduce complexity. It is a common phenomenon to research what 
seems like a simple R question and get 4-5 different suggestions about what you 
might need to do, including downloading specialized packages. Sometimes, these 
downloads have dependencies that are deprecated, out of date or non existent. 

More recently, responding to R studio's innovations, a 'data science' framework 
has been developed that offers an environment more similar to that used for 
clinical trials reporting. This framework, described in Hadley Wickham's R 
for Data Science book, free and online at  [https://r4ds.had.co.nz/].  This 
curriculum is based on the packages defined in that work. These include: DPLYR, 
TIDYVERSE,GGPLOT, LUBRIDATE , among others.

To investigate, you can enter installed.packages() at the console in RStudio. 
This function call returns a data structure ( a matrix,or an array ) to the screen. 

The 'packages' pane in the R Studio Gui also provides extensive help about each 
package that is installed, including examples and vignettes about using the package
and it's functionality. 

```{r "install_package",results='asis'}

## What type -- or class--of object is returned when you invoke the installed.packages function?

class(installed.packages())

## returns dimensions of the array, in Row,Column format
dim(installed.packages())

## or, the installed.packages function can be passed to the print function. 
## look at the 120th installed package, all columns
print(installed.packages()[120,])

##100th package, columns 1,2 and 3
print(installed.packages()[100,1:3])

## Since the detail is extensive, we can work with this new object to find the names of the columns. 
## here, we are looking at the contents of column 1 for all rows.
print(installed.packages()[,1])

#In addition to the print() function, R provides some other functions for looking 
#at what is in an R object. The View() function is particularly useful, it creates 
#a temporary spreadsheet which is navigable inside R Studio. Try changing 'print' 
#to 'View' in the above examples, entering the text at the R Studio Console. Note
#the capital 'V' in the word View. R is case sensitive.

```
## R libraries 

An R package consists of one or more packages or  code libraries. For example 
the 'tidyverse' package contains packages used in everyday data science work: 
dplyr, ggplot2, tidyr, readr,purrr,forcats, amd stringr.  WIthin each of these
packages, one can encounter one of more functions. For example, the dplyr package
has 'verbs' that work with data frames. These verbs can be chained together  
into a pipeline by %>% and are able to work with grouped data. They include 
functions like  arrange(), select(), mutate(), summarise() and slice(). 
These functions can be invoked individually or by a library() function call.

```{r "library_vs_packages"}
# to learn more about a package, the vignette function is helpful
# Issue this commend in the RStudio console: vignette("dplyr")

# It is also possible to refer to a specific function by applying 
#<<package::function*()>> syntax as in , DPLYR::filter() 

#filter() allows you to select a subset of rows in a data frame. Like all 
#single verbs, the first argument is the tibble (or data frame). The second and 
#subsequent arguments refer to variables within that data frame, selecting rows 
#where the expression is TRUE.

#For example, we can invoke the dplyr package as a code library  to select all 
# star wars characters with  yellow eyes:

dplyr::filter(dplyr::starwars, eye_color == "yellow")

# there is a less error-prone way. Here' we also select a gold-skinned, 
# yellow-eyed character, of which there should only be one:
# the library() function avoids having to use the '::' syntax.
library(dplyr)
starwars %>% filter(skin_color == "gold", eye_color == "yellow")

```
***
********************************************************************************
A library function call invokes the package from the group of installed packages
available for the R Version for use in an RStudio Session. These packages are 
installed in the R Studio space reserved for the user on a client or the desktop.   
The R version in use determines where these are updated on the user's machine. 
Once a package is installed, you pass the name to the library() function. If you 
have installed multiple versions of R, things can get messy. If in doubt, these 
versions can be deleted and reinstalled.

Later in this lesson, we will install a few utility packages from CRAN. Before 
installing any R packages, it is advisable to run the update.packages() function 
so that the packages are consistent with the version coming from the CRAN 
repository.

```{r "update_packages"}

## R may issue a message when you try to update packages in MS Windows 10:: 
#'lib = "C:/Program Files/R/R-4.1.0/library"' is not writable. . The workaround 
#is to navigate to the library in #Windows and unset the read-only flag as 
#administrator.
#Enter the following command without the comment '##' at the R STudio console 
#and press Enter. You may be prompted to #agree to update each package.

##update.packages("tidyr",ask=TRUE)

```

***


```{r "install_packages"}

## Create a character vector to pass to the install.packages() function. Names 
#should be quoted and are case-sensitive!

#Type the following command without the '##'comment at the R Studio console and 
#press Enter. You may be prompted to agree to update each package. 

## install.packages(c("here","janitor"))

```

***

## Installing packages from the R Studio Interface

The R Studio IDE provides a Packages pane. It is possible to install packages 
that way as well. You should try it by install.packages("tidyverse"). Tidyverse 
is a data science analytical package that creates data Structures similar to SAS 
datasets used in clinical trials.

***

## Tidyverse


```{r "library"}
## R libraries refer to physical locations on the user's R setup where these 
#packages and their base dependencies are installed. The R version in use 
#determines where these are updated on the user's machine. Once a package is 
#installed, you pass the anme to the library() function so you can use it.
##Load the tidyverse data analysis package you just installed:

library (tidyverse)

```


