---
title: "Creating a dataframe from SAS datasets"
output:
  html_document:
    theme: !expr sample(c("yeti", "united", "lumen"), 1)
    toc: yes
    toc_float: yes
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

## R Studio is designed to read data from the local file system or from an accessible external file system. 

## tidyverse, developed by R Studio, includes the HAVEN package to read SAS/SPSS/Stata files. This incorporates earlier work from the readr package.

## see also: https://tutorials.methodsconsultants.com/posts/reading-sas-spss-or-stata-files-into-r-using-haven/
```
***

## Objectives

After completing this unit, programmers should be able to:

- (1)  know to to list, access and load R Studio's built-in datasets
- (2)  access an internal location on their machine containing data and import it into an R data frame
- (3)  access an external location containing data and import it into an R data frame
- (4)  use these R data frames and write them as sas Datasets to their work environment
- (5)  compare the two datasets using SAS' proc compare to identify issues arising during processing
- (6)  identify the key characteristics of a r dataframe , and r tibble and ways in which these differ from SAS
datasets


***
 
## Internal data locations

- Built-ins:  There are several built-in internal data frames made available as part of the R Studio implementation. These are used for training purposes in the R Studio documentation and represent the universe of data that R users have contributed over the years. These built-ins provide a useful introduction to various data ranging from format categories to data frames. depending on the version of R.  
 
By default, these are written to a default location on the users' C: drive, during initial setup of RStudio, e.g.  "C:/Program Files/R/R-4.0.4/library" .


```{r "freebies", include=TRUE}
# To see the list of pre-loaded data, type the function data() at the R Studio console:

data()

## Or do it this way
data(package = .packages(all.available = TRUE))
#  Loading a built-in R data.

# Loading
data(mtcars)

# Print the first 6 rows
head(mtcars, 6)

```
***

- Local project data: R Studio can be set up to work with 'projects.'  These projects are typically stored somewhere on the user's C:\ drive , but can be configured to an accessible location external to the user's machine.  For example, I have copied several *.csv files into 'data\csv' contained in a project I set up with R studio. This is the path on my machine: 'C:\Users\thomasb\Documents\R\training\. ' A utility named "here" can point to this default location for the project and saves some typing. 

```{r "haven csv"}

library(tidyverse)
#the here package to tell R where your data is,
#install.packages("here")
library(here)

## read in a *.csv file into a tibble  with large/medium/small columns for each of 7 items
frames_wide<-read_csv(here("data/csv","frames_wide.csv"))
head(frames_wide)
```

- In a similar fashion, the sas subfolder I created contains *.xpt (sas Transport) files.  The haven package contained in the 'tidyverse' can read these into a tibble (a.k.a 'data frame').  

```{r "haven xpt"}

# Read in a sas *.xpt dataset. 
## Currently accessing *.xpt can be done only by a script:

library(haven)
 ae_xpt<-read_xpt(here("data/sas","ae.xpt"))
head(ae_xpt)

## from Navitas' SharePoint location:  
## OneDrive setup C:\Users\thomasb\Navitas Life Sciences\NDS R Training - Documents\General\Data
## note escaped '\' characters.
cm_xpt<-read_xpt(here("data/sas","cm.xpt"))
head(cm_xpt)

```
***

To load SAS *.sas7bdat datasets into R:
 - *.sas7bdat files can be imported from the file Menu in R Studio
 - or you can use datasetName<-haven::read_sas(data, path) in a script

```{r "haven sas7bdat"}
## reading in a *.sas7bdat is possible either way:
library(haven)
skinproduct_vfdemo <- read_sas("data/sas/skinproduct_vfdemo.sas7bdat",  NULL)
head(skinproduct_vfdemo)

```

##  Attaching, finding, summarizing
- - attach() utility reduces the need to use a$b referencing

## Some useful utilities for looking at SAS data in R

```{r "categorical attaching"}
require(utils)

attach (cm_xpt)

## Janitor package useful for character data
## NOT run install.packages("janitor")
library(janitor)
tabyl( CMDECOD,show_na=TRUE)

## turn off the attachment
detach(cm_xpt)
```

```{r "continuous attach"}

require(utils)

summary(women$height)   # refers to variable 'height' in the data frame
attach(women)
summary(height)         # The same variable now available by name
height <- height*2.54   # Don't do this. It creates a new variable
                        # in the user's workspace
find("height")
summary(height)         # The new variable in the workspace
rm(height)
summary(height)         # The original variable.
height <<- height*25.4  # Change the copy in the attached environment
find("height")
summary(height)         # The changed copy
detach("women")
summary(women$height)   # unchanged

```
***