---
title: "Unit3: cleaning data"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
  slidy_presentation: default
  ioslides_presentation: default
  word_document:
    toc: yes  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message=FALSE)
uid<-rstudioapi::userIdentity()
filePath<-file.path("C:/Users/",uid,"/Navitas Life Sciences/NDS R Training - Documents/General/Curriculum", fsep='/')
print(filePath)
setwd(filePath)

```
## Objectives

#### After completing this unit, R Studio users should :

- 1.1. Understand how to clean up column names manually using rename()
- 1.2  Understand how to work in the 'tidyverse'
- 1.3  Understand how to deal with Missing data
- 1.4  Know how to Modify and select columns from a data frame
- 1.5  Understand how to select rows from a data frame
- 1.5. Be able to “pipe” %>% data 
- 1.6  be able to use mutate()  to create or modify variables
- 1.7  Manipulate rows in a data frame

***
##  Attaching, finding, summarizing
- - attach(), reduces need to use a$b referencing

```{r "attach"}
require(utils)

summary(women$height)   # refers to variable 'height' in the data frame
attach(women)
summary(height)         # The same variable now available by name
height <- height*2.54   # Don't do this. It creates a new variable
                        # in the user's workspace
find("height")
summary(height)         # The new variable in the workspace
rm(height)
summary(height)         # The original variable.
height <<- height*25.4  # Change the copy in the attached environment
find("height")
summary(height)         # The changed copy
detach("women")
summary(women$height)   # unchanged
```
***
## Cross Tabulation and Table Creation
- - table() function:
- - tabulate() function -- see ?tabulate
--  Factors: categorical classes
- - table uses the cross-classifying factors to build a contingency table of the counts at each combination of factor levels.

```{r "table"}

require(stats) # for rpois and xtabs

## Simple frequency distribution
table(rpois(100, 5))

## Check the design:
with(warpbreaks, table(wool, tension))

## basic cross tab
table(state.division, state.region)

# simple two-way contingency table
with(airquality, table(cut(Temp, quantile(Temp)), Month))
a <- letters[1:3]
table(a, sample(a))                    # dnn is c("a", "")
table(a, sample(a), deparse.level = 0) # dnn is c("", "")
table(a, sample(a), deparse.level = 2) # dnn is c("a", "sample(a)")

## Confirming the cross tab:
## xtabs() <-> as.data.frame.table() :
UCBAdmissions ## already a contingency table
DF <- as.data.frame(UCBAdmissions)

class(tab <- xtabs(Freq ~ ., DF)) # xtabs & table
## tab *is* "the same" as the original table:
all(tab == UCBAdmissions)

all.equal(dimnames(tab), dimnames(UCBAdmissions))

a <- rep(c(NA, 1/0:3), 10)
table(a)                 # does not report NA's
table(a, exclude = NULL) # reports NA's

b <- factor(rep(c("A","B","C"), 10))
table(b)
table(b, exclude = "B")
d <- factor(rep(c("A","B","C"), 10), levels = c("A","B","C","D","E"))
table(d, exclude = "B")
print(table(b, d), zero.print = ".")

## NA counting:
is.na(d) <- 3:4
d. <- addNA(d)
d.[1:7]
table(d.) # ", exclude = NULL" is not needed
## i.e., if you want to count the NA's of 'd', use
table(d, useNA = "ifany")
```
***
```{r arrange}

```
***
```{r filter}

```
***

```{r group_by}

```
***
```{r summarize}

```
***